From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-04-17 Articles/2020-ClassicalMechanicsPhil/Submissions/BJPS/classical-phil-article-BJPS.tex
New Folders for BJPS, and cut parts of Section 5
Cut 5.1.1, 5.1.2, and 5.2. Bridged section transitions and checked for no ghost label calls

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Despite being coordinate-invariant, one cannot make the density truly coordinate independent. We therefore argue that some quantities are physically significant despite being coordinate-dependent. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we consider both the gauge redundancy of the conjugate momentum (Section~\ref{fundamental}) and of the Lagrangian (Section~\ref{gauge}).

@@

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. Differences in fundamentality are primarily differences in generality: more fundamental frameworks require fewer conditions to hold.\footnote{We also embrace related non-metaphysical uses of ``fundamental," discussed in Section~\ref{fundamental}.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the ``real" fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure as part of the theory's ontology, but as a formal device for encoding physical information or as a calculational tool. As Wilson illustrates, we can interpret classical mechanics while avoiding---and even rejecting---the search for what is `fundamental in the bottom-layer sense' \parencites*[53]{Wilson}.} Whether Hamiltonian phase space structure is metaphysically fundamental (in a classical world) does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically more fundamental, and we will argue for its epistemic and conceptual primacy as well. Indeed, our argument does not privilege any one of the various mathematical formalizations of Hamiltonian mechanics.\footnote{These include formulating Hamiltonian mechanics on a $2n+1$-dimensional contact manifold, on an extended phase space, on a Poisson manifold, and the standard formulation on a symplectic manifold that we typically use below.} Each of these frameworks has different pragmatic costs and benefits. Hence, it is plausible that there may be no single, uniquely necessary mathematical structure for mechanics, but rather a family of related structures. In other words, there may be no canonical minimal structure for formalizing the relevant physical models and physical assumptions.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics supplies a further reason for viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry" \parencites*[46]{Teh}.}  Additionally, each of these formalisms references some ``unphysical" coordinates or representational means, whose physical significance is borne out by their fruitfulness in modeling. We return to the physical significance of certain coordinate choices in Section~\ref{fundamental}.
%The mathematical structure that you ``need" might not be unique. Neither one of these three formalizations of Hamiltonian mechanics is necessarily more ``fundamental" or joint-carving than the others. Nevertheless, there are compelling physical and pragmatic motivations for preferring the extended phase space formulation of Hamiltonian mechanics. In this formulation, we gain an invariant Hamiltonian that is the conjugate variable of the proper time, while the normal Hamiltonian (which is not coordinate-invariant) still plays the role of generating time-translations.

@@

This completes our central argument for privileging Hamiltonian mechanics over Lagrangian mechanics. We have seen that Lagrangian systems are a proper subset of Hamiltonian systems, specifically those Hamiltonian systems that also satisfy the motion equivalence condition. In the remainder of this essay, we provide further reasons for privileging Hamiltonian mechanics. The next section describes how Hamiltonian mechanics provides the natural setting for the invariant density, which characterizes how a classical object's parts are distributed. Section~\ref{fundamental} argues that the invariant phase space density is more fundamental than the state variables, since conjugate momentum is a gauge-dependent quantity. In Section~\ref{Lagrangian} we provide a geometric interpretation of the principle of stationary action within Hamiltonian mechanics. Our construction makes manifest the gauge-dependency of the Lagrangian, providing a further reason to privilege the Hamiltonian. Finally, we respond to Curiel's arguments against the fundamentality of Hamiltonian mechanics.

@@

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues that since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position thereby undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms.\footnote{For arguments in favor of representing physical systems and explanations in exclusively intrinsic terms, see \textcites[]{Field,North,Chen2018}.} Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than the choice of state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics. The syntactic symmetry between position and momentum in the mathematical formalism does not entail that these quantities are equally physically significant. 
%\footnote{This illustrates how one can be mislead by the syntactical form of the mathematics to draw an interpretive conclusion that is not borne out by the physics.}

@@

The Hamiltonian formalism fixes this issue by assigning areas to regions in a way that is independent of the choice of unit-variables. To do so, it forces us to restrict ourselves to canonical coordinates, which explains why momentum has to change covariantly. The result is that we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in (in the sense of space-time coordinates). Any and all (time-independent) canonical coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See Section~\ref{gauge} for discussion.} In contrast, in the Lagrangian framework, there are at best special coordinate systems that allow for point-wise comparison of the density.
% % Try to precisely gloss/specify the condition for a coordinate system to be special in this way, such that it supports point--wise comparison. (Seems inappropriate to mention the Jacobian, since this matters for transforming from one coordinate system to another.)

In response to this deficiency of Lagrangian mechanics, one might wonder if it is possible to construct a coordinate-independent density. Section~\ref{independence} considers two strategies for doing so, showing why both fail. The first appeals to volume forms from differential geometry. The second appeals to the canonical volume form of Riemannian geometry, which suffices for the treatment of many Lagrangian systems. In each case, we will see that the coordinate-dependence of the physically significant density $\rho$ is unavoidable. Due to this coordinate-dependence, the density is in general not coordinate-invariant. Readers satisfied with the discussion so far can proceed to Section~\ref{comparing}, which illustrates one interpretive benefit of focusing on the density: it provides a better understanding of the relationship between Hamiltonian and Lagrangian mechanics within the hyper-regular domain.

\subsubsection{A coordinate-independent density?}
\label{independence}

In differential topology, a different mathematical object is called the `density,' namely a volume form (i.e. an alternating function from $n$ vectors to a real number) that is unaffected by changes in orientation.\footnote{The differential density generalizes volume forms in the case of a non-orientable manifold. However, we focus only on cases where the manifold is orientable. To simplify the exposition, we elide subtle differences between volume forms and differential densities, concentrating on the former.} Our physically significant density $\rho$ is actually the component of a volume form $\nu = \rho(x^i)  \epsilon dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $\epsilon$ is $1$ or $-1$ depending on the orientation. Note that $\nu$ combines the density and the volume into a coordinate-free object. Considering the total mass in the region $U$, we have $m(U) = \int \nu$ so essentially $\nu = dm$, i.e. the infinitesimal mass contribution we are integrating. But this means that the value returned by $\nu$ depends not only on how much stuff there is at a given point, but also on the size of the region identified by the $n$ vectors given to the form. In other words, $\nu$ has dimensions of only mass and is defined as a function of both a point and a volume element. In contrast, the density $\rho$ is just a function of the coordinates at a point, not the volume element, and has dimensions of mass over units. This shows that the coordinate-independent function $\nu$ cannot play the role of the coordinate-dependent density $\rho$.

In Riemannian geometry, we have a canonical volume form $dV = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $|g_{ab}|$ is the determinant of the metric tensor as expressed in local coordinates. In component notation, we can express the volume form as $\sqrt{|g_{ab}|}\epsilon_{a_1 ... a_n}$, where $\epsilon_{a_1 ... a_n}$ is the Levi-Civita symbol. This gives us an objective way to assign volumes to regions of space. We can then express any differential volume form $\nu = \hat{\rho} dV$ as the product of the volume form times a scalar function $\hat{\rho} \equiv \rho \ \epsilon \sqrt{|g_{ab}|}$. At first glance, this expression misleadingly suggests that $\hat{\rho}$ is a coordinate-independent density: it is a scalar quantity with the correct dimensions of mass over volume, meeting our standard criteria for a density. However, note that the numerical value of $\hat{\rho} $ equals the value of $\rho$ only if $|g_{ab}| = 1$, where $1$ is a pure number. This match happens only in special coordinate frames and, for this reason, $\hat{\rho} $ remains coordinate-dependent. In other words, the metric tensor defines a unit for the volume form (e.g. meters-cubed), and we always express the volume in those units, regardless of what coordinates we chose. This would be a natural choice of volume in a Cartesian system, for example, but not in a spherical one. 
%That is: to This is only possible if the determinant of $g_{ab}$ equals one. Hence, we must define $\hat{\rho} $ in a special coordinate system, namely one where $|g_{ab}| = 1$. 
%det g_ab =1 means cells you construct are equal volume. whereas cartesian means not just equal volume but sides (like cubes)
%Fortunately, we can always find a set of coordinates that, at least locally, makes $|g_{ab}|$ constant. Since the square root of the determinant $|g_{ab}|$ is the Jacobian of the transformation to those coordinates, $|g_{ab}|$ will be constant only when this Jacobian equals one.

For example, the metric tensor of a two dimensional space may define length in terms of meters. The canonical volume form would then define areas in meters squared. This would be the natural choice of volume unit in Cartesian coordinates, where in fact $|g_{ab}| = 1$ and the volume form $dV = dx \wedge dy$. In polar coordinates, the natural choice of units would be meter $\cdot$ radians. However, the volume form will still return meters-squared, since we will have $|g_{ab}| = r^2$ and $dV = r dr \wedge d\theta$. This means that $\hat{\rho}$ at best expresses the density in a particularly convenient Cartesian-like coordinate system: one in which the volume is expressed in special units and we can compare the density point-wise.\footnote{This is similar to what happens with the concept of proper time: it is defined in the co-moving frame, so it is coordinate-dependent. Yet, the expression $d\tau = \frac{1}{c} \sqrt{g_{ab} dx^a dx^b}$ is coordinate-invariant. This coordinate-invariance derives from the fact that we have conventionally agreed to define proper time in terms of the object's co-moving frame, rather than in terms of a different frame.} By conventionally agreeing to work in a coordinate system where $|g_{ab}| = 1$, we can make the density $\hat{\rho}$ coordinate-invariant. But, the property of point-wise comparison remains coordinate-dependent: it requires working in a particular coordinate system. Outside the context of Hamiltonian mechanics, the best we can do is obtain the property of point-wise comparison in a special coordinate system.
% General idea from one-11 meeting: in the general case, in order to make the density a coordinate-invariant [so that we all can agree on its value], we have to make a global choice of coordinate system. We need to choose coordinate such that the determinant of the metric tensor is equal to 1. So we are seemingly saying that there are coordinate systems that are special, namely those where we can do point-wise comparison. So this poses a problem for the view that nature does not privilege one coordinate system over another. Possible response from the metaphysicians: perhaps the real fundamental physical object is a quotiented-out version of the density.

% From before 1-11 : Again, the upshot here is that we cannot avoid using a coordinate-dependent density. 
%We can also see this if we look at the units of $\hat{\rho}$, which are in terms of inverse volume as expressed in the privileged system. From 1-4 meeting: I thought that $\hat{\rho}$ has units of mass over volume.
% Original bad explanation (see 1-11-21 meeting, around 11:13 am): Nevertheless, $\hat{\rho}$ remains coordinate-dependent because it is a function of $|g_{ab}|$, which depends on the choice of coordinates and is, in general, a function of the point.

\subsubsection{Comparing densities on the tangent vs. cotangent bundles}
\label{comparing}

% % Wonder if it is possible to move this material to the `interpreting Lagrangian mechanics' section?

Focusing on the density has the further benefit of elucidating the relationship between Hamiltonian and Lagrangian mechanics, especially within the hyper-regular domain. Consider a distribution over position and velocity, i.e. on the tangent bundle in Lagrangian mechanics. Note that the velocities transform in the same way (i.e. contravariantly) as differentials in position. Since both position and velocity transform contravariantly, the density will transform according to $\rho(x^i, v^j) = \sqrt{|g_{ab}|} \sqrt{|g_{ab}|} \rho(\hat{x}^i, \hat{v}^j) = |g_{ab}| \rho(\hat{x}^i, \hat{v}^j)$. The canonical volume form for the tangent bundle becomes: $dV_{TM} = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n \bigwedge \epsilon \sqrt{|g_{ab}|} dv^1 \wedge dv^2 \wedge ... \wedge dv^n = |g_{ab}| \bigwedge dx^i \bigwedge dv^j$. Hence, the determinant of the metric tensor equals the component of the volume form of the tangent bundle. The relationship to the density as expressed in phase space is therefore the following:
\begin{equation}
	\rho(q^i, p_j) = m |g_{ab}| \rho(x^i, v^j).
\end{equation}
A change of unit-variables will induce a contravariant change in the velocity and a covariant change in the momentum. The left side will remain invariant. The two terms $|g_{ab}|$ and  $\rho(x^i, v^j)$ on the right side will change in the opposite way so that their product is invariant. Focusing on densities, then, brings about a physically-motivated way to understand the relationship between the cotangent and the tangent bundle, where the metric tensor plays a role in defining the geometry of the latter similar to how the symplectic form defines the geometry of the former: it defines the count of states so that densities can be properly tracked.

These relationships also allow us to better understand the hyper-regular case, and in particular the case of a massive particle. Note that, in general, the relationship between densities over the cotangent and tangent bundles is, as we saw:
\begin{equation} \label{densities_relationship}
\rho(q^i, p_j) = \left|\frac{\partial v^j}{\partial p_k}\right| \rho(x^i, v^j)  = \left|\frac{\partial H}{\partial p_k \partial p_j}\right| \rho(x^i, v^j).
\end{equation}
In the case of massive particles, where the Hamiltonian (and the Lagrangian) are quadratic, the Hamiltonian of Equation~\ref{massive_Hamiltonian} entails that $\frac{\partial H}{\partial p_k \partial p_j} = m g^{kj}$. This is the case where the densities in the tangent bundle can be properly understood in terms of Riemannian geometry. For the more general case, we note that
\begin{equation}
\begin{aligned}
\delta^i_j = \frac{\partial v^i}{\partial v_j} = \frac{\partial v^i}{\partial p_k}\frac{\partial p_k}{\partial v_j} &= \frac{\partial}{\partial p_k}\left(\frac{\partial H}{\partial p_i}\right) \frac{\partial }{\partial v_j} \left( \frac{\partial L}{\partial v_k} \right) \\
\Rightarrow \frac{\partial L}{\partial v^k \partial v^j} &= \left[ \frac{\partial H}{\partial p_k \partial p_j} \right]^{-1}
\end{aligned}
\end{equation}
which means that the Hessian of the Hamiltonian is the inverse of the Hessian of the Lagrangian, and so are their determinants. Recall that the hyper-regular case corresponds to the case where the determinant of the Hessian of the Lagrangian is non-zero. Hence, Equation~\ref{densities_relationship} tells us that in this case, we can express the density of states over position and velocity.

% Remember: the density is coordinate-invariant but not coordinate-independent; it is actually coordinate-dependent (i.e. it does not admit of something like a coordinate-free definition)
% Also remember: there are quantities that are both coordinate-variant and coordinate-dependent that are still physically significant, e.g. choices of q's and p's (?)
% Coordinate-invariant in general means that definition of the quantity involves referencing coordinates, but if you change the coordinates, then the value of the quantity remains the same. (E.g. transforms as a scalar). so densities are never coordinate-independent. defining a density requires a choice of units, so involves referencing units. 
%see 57.3 minutes of 12-21 recording: when we define the density originally, it is coordinate-free (rho as a function of the state space of the particles, S). But when we go to express it numerically, then it becomes coordinate-dependent. (I.e. when we express it over a manifold, then it needs to be a density for integration, and hence it becomes coordinate-dependent). 
% see 51 minutes of 12-21 recording: Densities (the coordinate-dependent kind) are physically significant partly because they tell you how much stuff is in each state

%\subsection{Fundamental density and gauge-dependent momentum}
\subsection{Coordinate-dependent yet fundamental} %; Coordinate-variant yet significant
\label{fundamental}

Section on Ham. Not everything physical has to be coordinate independent or invariant.

* In phase space, we do have coordinate invariance and point-wise comparison.  Momentum changes covariantly. It is the density that is truly fundamental, not the coordinates in general. Though, the q's are priviledged from epistemic access. The p's are not, in fact we can define gauge transformation as things that keep the q's the same but change the p.

*[Incorporate the following from section 5.2] Not all ``state variables" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a gauge quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. $->$ Invariant density is really the important physical element/object in Hamiltonian mechanics.

* Units are physically significant. In fact, the ability to express the densities in units is what gives us the symplectic structure. It's a physical requirement that we must be able to express the density in some unit. [Sort of transcendental/regulative assumption]
* Invariant densities: physical significance connected to coordinate invariance. [incorporated]
* Density comparison is independent of coordinates. We can compare densities point-wise without worry. [incorporated]

* Makes the fundamental role that phase space plays in statistical mechanics more physically salient/perspicuous. Areas/volumes are measures of number of states. They are coordinate invariant. Also the Shannon/Gibbs entropy are coordinate invariant.

% Section~\ref{action} uses the invariant density to provide a geometrical interpretation of the action principle.

As argued at the start of Section~\ref{density}, the invariant density is physically fundamental because it specifies the state of the classical system. Yet, as Section~\ref{dependence} explained, the density is always coordinate-dependent, even when coordinate-invariant. In other words, there is no coordinate-independent (or coordinate-free) definition of the invariant phase space density. Affording the density fundamental significance therefore places us at odds with those who take coordinate-independence as a necessary condition for fundamentality.

%We have just shown that Hamiltonian mechanics provides the proper setting for representing the density. By representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) canonical coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See Section~\ref{gauge} for discussion.} 
%Collectively, these considerations support interpreting the invariant phase space density as not only physically significant but also fundamental. 
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.

Proponents of this necessary condition might respond by noting that although the density $\rho_{c, \xi}$ is not coordinate-independent, the distribution $\rho_{c}$ is. For the distribution is equivalent to the classical state, which is a thing in the world, i.e. an ontic entity. We grant that the distribution is ontically fundamental in a sense that does not accrue to the density (which represents the distribution). However, we maintain that the density itself is of fundamental significance, both epistemically and physically. For it is only through the density $\rho_{c, \xi}$ that we gain epistemic access to the distribution. Although the choice of coordinates $\xi$ is fairly arbitrary, we need to make some choice in order to quantitatively measure the distribution $\rho_{c}$ through experiments. Additionally, defining the density is necessary for deriving symplectic structure, as shown in Section~\ref{infinitesimal}. Insofar as our opponent views the symplectic structure of Hamiltonian mechanics as fundamental, they must view the density as equally fundamental---despite its coordinate-dependence. For the same reason, insofar as our opponent refuses to grant the density fundamental status, they will have to withhold this same status from the symplectic structure. Hence, the coordinate-independent notion of fundamentality needs to be supplemented with a coordinate-dependent, epistemic notion of fundamentality. This latter notion is no less important, and it shows that it is a mistake to desire a completely ``intrinsic" formulation of a physical theory.
% % Further claim: the ontic, coordinate-independent notion of fundamentality is insufficient for doing physics. We need a corresponding epistemic notion of fundamentality, since we units, state variables, and the density in order to connect the distribution to experiment. One can argue that any particular choice of state variables or units or coordinates for the density is arbitrary, but we need some arbitrary choice in order to quantitatively measure the distribution through experiments.
% % Someone might also worry that the density cannot be more fundamental than the state variables, since we define the density over state variables. In response, we can say the following: whereas the density is coordinate-invariant, the state variables are not. So the state variables are more arbitrary than the density. Note that we do not directly measure states of particles. We measure quantities. So the only way to measure the distribution is to express it as a density in terms of some quantities, i.e. in terms of some state variables. So we need the state variables to give epistemic access to the distribution, via the density. So the state variables themselves are not important (e.g. we can choose the unit basically however we want, and the conjugate momentum has even more freedom, due to gauge redundancy). The physically important thing is access to the distribution, which we gain through the density. The classical system is made up of pieces and the density lets us say how many pieces we have in various states.
% A defender of the necessity of coordinate-independence for fundamentality might respond as follows:

% % [To do: provide literature references for this and motivating quotations; likewise for earlier remark about intrinsicality: Field, North, Chen]

%[XXX To do: insert discussion of statistical mechanics and entropy here: statistical mechanics works best on phase space because phase space gives us a uniform measure over q and p. Focusing on the density makes the fundamental role that phase space plays in statistical mechanics more salient/perspicuous. Areas/volumes are measures of the number of states. They are coordinate invariant. Also the Shannon and Gibbs entropies are coordinate invariant.]\footnote{For further discussion of the importance of phase space for statistical mechanics and entropy, see \textcites{shpmp} [omit this for review?].}

Instead, of coordinate-independence, we take coordinate-invariance as a weaker necessary (but not sufficient) condition for fundamental physical significance. Being necessary for fundamental significance does not, however, make coordinate-invariance necessary for \textit{non-fundamental} physical significance. Many representational choices are coordinate-variant yet still physically significant.\footnote{Indeed, many physically significant quantities are gauge-dependent, such as conjugate momentum and the Lagrangian (see Sections~\ref{gauge_momentum} and~\ref{gauge}). Insofar as gauge-dependency is analogous to being coordinate-variant, this supplies a further reason for denying that coordinate-invariance is a necessary condition for physical significance.} Our position here again places us at odds with others in the literature. For instance, North claims that coordinate choices are ``merely arbitrary choices of description" \parencites*[61]{North}. However, some choices of coordinates are particularly well-suited to certain problems, such as the use of action--angle variables for integrable systems, or spherical coordinates for systems with spherical symmetry. Similarly, we saw in Section~\ref{dependence} that some coordinate choices in Lagrangian mechanics are physically significant because they enable the property of point-wise comparison. This is related to the fact that, in Lagrangian mechanics, Cartesian coordinates are a physically significant coordinate choice because they decompose position and velocity variables as independent degrees of freedom. This independence does not hold generically across coordinate systems.\footnote{In contrast, canonical position and conjugate momentum variables always characterize separable/independent degrees of freedom in the Hamiltonian framework.} In a similar vein, the next section shows that the conjugate momentum is gauge-dependent, but it is presumably still physically significant. 
%Realized that the following remarks probably go with earlier argument about coordinate-dependence: Finally, as the derivation in Section~\ref{infinitesimal} shows, a necessary condition for defining the symplectic structure is that we can express the density $\rho_{c, \xi}$ in a particular choice of units. If we did not have the physical ability to define the density in some units we would not be able to define this structure. This provides another instance of a physically significant but coordinate-dependent representational choice.

% % Could insert a summary paragraph either here or at the end of the next subsection explaining the following: coordinate-invariance and gauge-invariance are both necessary for fundamental physical status. But they're not necessary for being physically significant. Both coordinate-variant and gauge-dependent quantities can still be physically significant, although they cannot be physically fundamental. So in this way, we view gauge-dependency as a kind of coordinate-variance (rather than a kind of coordinate-dependence). And gauge-invariance is analogous to a kind of coordinate-invariance, rather than a kind of coordinate-independence. And perhaps this helps explain the physical significance of gauge-variant quantities. 
% Question: is any quantity that is coordinate-variant automatically coordinate-dependent? If so, then is coordinate-variance a special case of coordinate-dependence? And so gauge-variance would also be a special case of coordinate-dependence?

@@

Additionally, the gauge-dependency of conjugate momentum shows that all state variables are \textit{not} on equal physical footing. Although the canonical position and momentum variables are on equal syntactical footing within symplectic mechanics, this does not entail that they are equally physically significant. To see this, recall that the phase space density remains invariant upon changing the unit-variables $q$, inducing a state variable transformation on the conjugate momenta $p$ (which compensate by transforming covariantly). It is in this sense that the unit-variables $q$---rather than the conjugate momenta $p$---define the units of physical measurement, allowing us to assign numbers to physical quantities.\footnote{As discussed in Section~\ref{fundamental}, these units are physically significant despite not being coordinate-invariant.} Consequently, although symplectic mechanics allows arbitrary mixings of $q$'s and $p$'s that preserve the symplectic manifold, not all of these combinations are of equal physical significance. In general, the canonical position coordinates $q$ are epistemically privileged because they are linked with measurements, in a sense that the logically-derivative $p$'s are not. Unlike both velocity and the kinetic momentum $m g_{ji} v^i$, conjugate momentum is not an experimentally measurable quantity, because its specification involves a gauge choice.

>>>>

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Despite being coordinate-invariant, one cannot make the density truly coordinate independent. We therefore argue that some quantities are physically significant despite being coordinate-dependent. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we consider both the gauge redundancy of the conjugate momentum (Section~\ref{gauge_momentum}) and of the Lagrangian (Section~\ref{gauge}).

@@

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. Differences in fundamentality are primarily differences in generality: more fundamental frameworks require fewer conditions to hold.\footnote{We also embrace related non-metaphysical uses of ``fundamental," discussed in Section~\ref{density}.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the ``real" fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure as part of the theory's ontology, but as a formal device for encoding physical information or as a calculational tool. As Wilson illustrates, we can interpret classical mechanics while avoiding---and even rejecting---the search for what is `fundamental in the bottom-layer sense' \parencites*[53]{Wilson}.} Whether Hamiltonian phase space structure is metaphysically fundamental (in a classical world) does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically more fundamental, and we will argue for its epistemic and conceptual primacy as well. Indeed, our argument does not privilege any one of the various mathematical formalizations of Hamiltonian mechanics.\footnote{These include formulating Hamiltonian mechanics on a $2n+1$-dimensional contact manifold, on an extended phase space, on a Poisson manifold, and the standard formulation on a symplectic manifold that we typically use below.} Each of these frameworks has different pragmatic costs and benefits. Hence, it is plausible that there may be no single, uniquely necessary mathematical structure for mechanics, but rather a family of related structures. In other words, there may be no canonical minimal structure for formalizing the relevant physical models and physical assumptions.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics supplies a further reason for viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry" \parencites*[46]{Teh}.}  Additionally, each of these formalisms references some ``unphysical" coordinates or representational means, whose physical significance is borne out by their fruitfulness in modeling.
% We return to the physical significance of certain coordinate choices in Section~\ref{fundamental}.
%The mathematical structure that you ``need" might not be unique. Neither one of these three formalizations of Hamiltonian mechanics is necessarily more ``fundamental" or joint-carving than the others. Nevertheless, there are compelling physical and pragmatic motivations for preferring the extended phase space formulation of Hamiltonian mechanics. In this formulation, we gain an invariant Hamiltonian that is the conjugate variable of the proper time, while the normal Hamiltonian (which is not coordinate-invariant) still plays the role of generating time-translations.

@@

This completes our central argument for privileging Hamiltonian mechanics over Lagrangian mechanics. We have seen that Lagrangian systems are a proper subset of Hamiltonian systems, specifically those Hamiltonian systems that also satisfy the motion equivalence condition. In the remainder of this essay, we provide further reasons for privileging Hamiltonian mechanics. The next section describes how Hamiltonian mechanics provides the natural setting for the invariant density, which characterizes how a classical object's parts are distributed. The invariant phase space density is more fundamental than the state variables, since conjugate momentum is a gauge-dependent quantity. In Section~\ref{Lagrangian} we provide a geometric interpretation of the principle of stationary action within Hamiltonian mechanics. Our construction makes manifest the gauge-dependency of the Lagrangian, providing a further reason to privilege the Hamiltonian. Finally, we respond to Curiel's arguments against the fundamentality of Hamiltonian mechanics.

@@

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than the choice of state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics. The syntactic symmetry between position and momentum in the mathematical formalism does not entail that these quantities are equally physically significant. 
%\footnote{This illustrates how one can be mislead by the syntactical form of the mathematics to draw an interpretive conclusion that is not borne out by the physics.}

@@

The Hamiltonian formalism fixes this issue by assigning areas to regions in a way that is independent of the choice of unit-variables. To do so, it forces us to restrict ourselves to canonical coordinates, which explains why momentum has to change covariantly. The result is that we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in (in the sense of space-time coordinates). Any and all (time-independent) canonical coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See Section~\ref{gauge} for discussion.} In contrast, in the Lagrangian framework, there are at best special coordinate systems that allow for point-wise comparison of the density.\footnote{One might wonder if it is possible to construct a coordinate-independent density, perhaps using either i) volume forms from differential geometry or ii) the canonical volume form of Riemannian geometry. Both of these strategies fail, although space precludes a demonstration.}

% % Try to precisely gloss/specify the condition for a coordinate system to be special in this way, such that it supports point--wise comparison. (Seems inappropriate to mention the Jacobian, since this matters for transforming from one coordinate system to another.)

@@

Additionally, the gauge-dependency of conjugate momentum shows that all state variables are \textit{not} on equal physical footing. Although the canonical position and momentum variables are on equal syntactical footing within symplectic mechanics, this does not entail that they are equally physically significant. To see this, recall that the phase space density remains invariant upon changing the unit-variables $q$, inducing a state variable transformation on the conjugate momenta $p$ (which compensate by transforming covariantly). It is in this sense that the unit-variables $q$---rather than the conjugate momenta $p$---define the units of physical measurement, allowing us to assign numbers to physical quantities. Consequently, although symplectic mechanics allows arbitrary mixings of $q$'s and $p$'s that preserve the symplectic manifold, not all of these combinations are of equal physical significance. In general, the canonical position coordinates $q$ are epistemically privileged because they are linked with measurements, in a sense that the logically-derivative $p$'s are not. Unlike both velocity and the kinetic momentum $m g_{ji} v^i$, conjugate momentum is not an experimentally measurable quantity, because its specification involves a gauge choice.
%\footnote{As discussed in Section~\ref{fundamental}, these units are physically significant despite not being coordinate-invariant.}
